---
title: "Gapminder-Exploration"
output: 
  html_document:
    keep_md: yes
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r cars}
library(gapminder)
library(tidyverse)
```

## Basic Exploration of Data

We can figure out what kind of dataset it is by using the following command:
```{r}
class(gapminder)
```
This dataset is therefore a data frame.

We extract the number of rows and columns in the dataset as follows:
```{r}
nrow(gapminder) # extracts number of dataset rows
ncol(gapminder) # extracts number of dataset columns
```
As we can see, the number of rows is 1704 and the number of columns is 6.

This gives us the types of all variables in the dataset:
```{r}
str(gapminder)
```

They're 6 variables, country, continent, year, lifeExp, pop, and gdpPercap. Their types are Factor, Factor, integer, number, interger, and number, respectively.

## Exploration of Individual Variables

I will be exploring the qualitative "continent" and the quantitative variable "lifeExp".
```{r}
summary(gapminder$continent)
summary(gapminder$lifeExp)
```
For the continent variable, they're 624 datapoints with Africa as a continent, 300 with the Americas, 396 with Asia, 360 with Europe, and 24 with Oceania. 
For the life expectancy variable, the maximum value is 82.60, the minimum value is 23.60, and the median value is 60.71. 

## Plotting and various dplyr Commands

For the following commands, I often use the piping syntax, to improve the readability of the code.

This is how we can select a column, in this case "lifeExp", from the dataset. We use the select, as follows:
```{r}
gapminder %>%
  select(lifeExp)
```

This is how we can select certain rows, in this case rows where the life expectancy exceeds 40, from the dataset. We use filter, as follows:
```{r}
gapminder %>%
  filter(lifeExp > 40)
```

The following code gives the mean life expectancy by year.
```{r}
gapminder %>%
  group_by(year) %>% # this line groups the data by year
  summarise(mean(lifeExp)) # this line compute the mean life expectancy for each year
```

As one would expect, global life expectancy continues to increase as time goes on.

The following code shows the max population across all countries for various years between 1972 to 2007.
```{r}
gapminder %>%
  filter(year >= 1972) %>%
  group_by(year) %>% # this line groups the data by year, if the year is greater than or equal to 1972
  summarise(max(pop)) 
```

The following commands plot life expectancy versus GDP per capita, in 1997, as a scatterplot.
```{r}
gapminder %>%
  filter(year == 1997) %>%
  ggplot(aes(gdpPercap, lifeExp)) + geom_point()
```

From the plot, it seems as though places with a higher GDP per capita have higher life expectancies in general.

The following commands plot a boxplot, comparing the GDP per capita of the various continents (where each data point is a country) in 2007.
```{r}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(continent, gdpPercap)) + geom_boxplot()
```

As we can see from the plot, it seems as though in general that Europe and Oceania have the highest GDP per capita, at least in 2007.

The following code produces a dotplot of the GDP per capita's of countries in 2007 (note that dotplot is very similar to a histogram, where each dot corresonds to a data point).
```{r}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(gdpPercap)) + geom_dotplot(dotsize=0.5, binwidth = 2000) # plots a dotplot with bins of size 2000
```

As we can see, most countries have GDP per capita's between 0 to 40000, with very few greater than that.
