---
title: "Gapminder-Exploration"
output: 
  html_document:
    keep_md: yes
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r cars}
library(gapminder)
library(tidyverse)
```

## Exploration of Data

We can figure out what kind of dataset is by using the following command:
```{r}
class(gapminder)
```
This dataset is therefore a data frame.

We extract the number of rows and columns in the dataset as follows:
```{r}
nrow(gapminder) # extracts number of dataset rows
ncol(gapminder) # extracts number of dataset columns
```
As we can see, the number of rows is 1704 and the number of columns is 6.

This gives us the types of all variables in the dataset:
```{r}
str(gapminder)
```

They're 6 variables, country, continent, year, lifeExp, pop, and gdpPercap. Their types are Factor, Factor, integer, number, interger, and number, respectively.

## Exploration of Individual Variables

I will be exploring the qualitative "continent" and the quantitative variable "lifeExp".
```{r}
summary(gapminder$continent)
summary(gapminder$lifeExp)
```
Describe results.

## Plotting and various dplyr Commands

For the following commands, I often use the piping syntax, to improve the readability of the code.

This is how we can select a column, in this case "lifeExp", from the dataset. We use the select, as follows:
```{r}
gapminder %>%
  select(lifeExp)
```

This is how we can select certain rows, in this case rows where the life expectancy exceeds 40, from the dataset. We use filter, as follows:
```{r}
gapminder %>%
  filter(lifeExp > 40)
```

The following code gives the mean life expectancy by year.
```{r}
gapminder %>%
  group_by(year) %>%
  summarise(mean(lifeExp)) 
```

As one would expect, global life expectancy continues to increase as time goes on.

The following code shows the max population across all countries every 5 years between 1972 to 2007.
```{r}
gapminder %>%
  filter(year >= 1972) %>%
  group_by(year) %>%
  summarise(max(pop)) 
```

The following commands plot life expectancy versus GDP per capita, in 1997, as a scatterplot.
```{r}
gapminder %>%
  filter(year == 1997) %>%
  ggplot(aes(gdpPercap, lifeExp)) + geom_point()
```

The following commands plot a boxplot, comparing the GDP per capita of the various continents (where each data point is a country) in 2007.
```{r}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(continent, gdpPercap)) + geom_boxplot()
```

As we can see from the plot, it seems like in general that Europe and Oceania have the highest GDP per capita, at least in 2007.
